LDpred2: better, faster, stronger

### 2020-05-31 paper/paper-ldpred2.tex
save

\abstract{	
Polygenic scores have become a central tool in human genetics research.
LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
However, LDpred has limitations that may result in limited predictive performance.
Here we present LDpred2, a new version of LDpred that addresses these issues.
We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns parameters from data.
We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
We then show that LDpred2 compares favorably to other polygenic score methods.
LDpred2 is implemented in R package bigsnpr.
}

@@

Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
Six simulation scenarios are used, each repeated 10 times.
In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
In these scenarios, infinitesimal models perform similarly. 
When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.7e-4 and 2.7e-3). 
For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2 [TODO: VERIF NEW NUMBERS].
In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in long-range LD. 
In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.

%% FIGURE 1 HERE
\begin{figure}[htbp]
\centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
\caption{Results of the six simulation scenarios. 
Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25.5-33.5 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
%See corresponding values in table S1.
}
\label{fig:AUC-simu}
\end{figure}

@@

Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1. These improvements are also significant for MDD and CAD, albeit to a lesser extend. For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 64.4\% with LDpred2, and from 54.9\% to 74.3\% for T1D. 
For Asthma and RA, predictive performances of LDpred1 and LDpred2 are similar.

As in simulations, the sparse version of LDpred2 performs equally well as the original version.
For PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly [TODO: CHANGE].

@@

When comparing to other methods, we only report the best LDpred2 model (chosen using the validation set). 
LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).

[TODO: COMPLETE]

@@

Here we present LDpred2, a new version of LDpred. 
LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
We have therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.

@@

In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
When comparing to recently derived methods that showed higher predictive performance as compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, which may break some of the assumptions behind their respective models.
Selection of variants is therefore a direction of future research for us.

[TALK ABOUT QUALITY CONTROL ON AF? -> ON SD, NEED A SUPP NOTE] 
[ALSO TALK ABOUT MODEL FOR BINARY DATA]

@@

We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 352,320 individuals as test set.
We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
We defined phenotypes as in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section). [TODO: TALK ABOUT QC HERE]

@@

In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}. For C+T, we choose the best set of parameters on each chromosome individually and combine them using a proper scaling as done for LDpred2 (see section \ref{sec:bychr}).
For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, \dots, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.

@@

%\begin{minipage}{.7\textwidth}
\begin{algorithm}[H]
\caption{LDpred, with hyper-parameters $p$ and $h^2$, LD matrix $\boldsymbol{R}$ and summary statistics $\boldsymbol{\hat{\gamma}}$, $\boldsymbol{\text{se}(\hat{\gamma})}$ and $\boldsymbol{n}$}\label{algo:ldpred}
\begin{algorithmic}[1]
	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}}$ \Comment{Initialization of scaled marginal effects (see previous section)}
	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$ \Comment{Initialization of posterior means}
	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
	\For {each variant $j$} \Comment{All variants}
	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
	\EndFor
	\If {$k > N_\text{burn-in}$} 
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} + \boldsymbol{\omega}$ 
	\EndIf
	\EndFor
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$ \Comment{Average of all $\boldsymbol{\omega}$ after burn-in}
	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}$ \Comment{Return posterior means, scaled back (see previous section)}
\end{algorithmic} 
\end{algorithm}
%\end{minipage}

%[CHANGE NOTATION OF SUMSTATS?]

@@

We run LDpred2 per chromosome.
Even if it is possible to run LDpred2 genome-wide, this approach has two limitations. First, it can be memory and computationally demanding to do so. For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory. 
Doubling to 2M variants would require 128 GB of RAM to store the matrix.
Second, as noted in \cite{prive2019making}, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results. 
To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC. 
Using the regression would also enable adjusting for covariates in this step.

@@

[TODO]

@@

>>>>

\abstract{	
Polygenic scores have become a central tool in human genetics research.
LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
However, LDpred has limitations that may result in limited predictive performance.
Here we present LDpred2, a new version of LDpred that addresses these issues.
We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns parameters from data.
We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
We then show that LDpred2 compares favorably to other polygenic score methods developed more recently.
LDpred2 is implemented in R package bigsnpr.
}

@@

Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
Six simulation scenarios are used, each repeated 10 times.
In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
In these scenarios, infinitesimal models perform similarly. 
When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.7e-4 and 2.7e-3). 
For example, in simulations with 300 causal variants, a mean AUC of 73.6\% is obtained with LDpred1 while a value of 81.7\% is obtained with LDpred2.
In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled from a long-range LD region. 
In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.

%% FIGURE 1 HERE
\begin{figure}[htb]
\centerline{\includegraphics[width=0.85\textwidth]{AUC-simu}}
\caption{Results of the six simulation scenarios. 
Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25.5-33.5 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
%See corresponding values in table S1.
}
\label{fig:AUC-simu}
\end{figure}

@@

Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1. These improvements are also significant for MDD and CAD, albeit to a lesser extend. 
For example, for BRCA, AUC improves from 58.9\% with LDpred1 to 64.6\% with LDpred2, and from 57.4\% to 77.0\% for T1D. 
For Asthma and RA, predictive performances of LDpred1 and LDpred2 are similar.
As in simulations, the sparse version of LDpred2 performs equally well as the original version.
Moreover, LDpred2-auto also perform  very well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set). 
LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).

@@

@@

Here we present LDpred2, a new version of LDpred. 
LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
We have therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.

@@

In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
When comparing to recently derived methods that showed higher predictive performance compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, or detrimental as it may break some of the assumptions behind their respective models.
Selection of variants is therefore a direction of future research for us.

@@

We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 352,320 individuals as test set.
We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
We defined phenotypes as in \cite{prive2019making}. Some quality control is applied to summary statistics (see Supplementary Note).
For more details, please refer to our R code (Software and code availability section).

@@

In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, \dots, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.

@@

\begin{algorithm}[H]
\caption{LDpred, with hyper-parameters $p$ and $h^2$, LD matrix $\boldsymbol{R}$ and summary statistics $\boldsymbol{\hat{\gamma}}$, $\boldsymbol{\text{se}(\hat{\gamma})}$ and $\boldsymbol{n}$}\label{algo:ldpred}
\begin{algorithmic}[1]
	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}}$ \Comment{Initialization of scaled marginal effects (see previous section)}
	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$ \Comment{Initialization of posterior means}
	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
	\For {each variant $j$} \Comment{All variants}
	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
	\EndFor
	\If {$k > N_\text{burn-in}$} 
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} + \boldsymbol{\omega}$ 
	\EndIf
	\EndFor
	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$ \Comment{Average of all $\boldsymbol{\omega}$ after burn-in}
	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}$ \Comment{Return posterior means, scaled back (see previous section)}
\end{algorithmic} 
\end{algorithm}

@@

We run LDpred2 per chromosome.
Even if it is possible to run LDpred2 genome-wide, this approach has two limitations. First, it can be memory and computationally demanding to do so. For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory. 
Doubling to 2M variants would require 128 GB of RAM to store the matrix.
Second, as noted in \cite{prive2019making}, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results. 
To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC.

@@

[TODO: EXPORT CODE FROM CLUSTER]

@@

\section*{Supplementary Note: quality control of summary statistics}

For summary statistics of binary traits derived from a logistic regression, 
\begin{equation}\label{eq:approx-sd}
\text{sd}(\boldsymbol{g_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
\end{equation}
where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. Anyway, we recommend to verify this assumption and to perform some quality control.
Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
However, in real data applications, where summary statistics comes from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set.
We removed variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.7\textwidth]{sd-approx-simu}}
\caption{In simulations, standard deviations derived from summary statistics based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set.}
\label{fig:sd-approx-simu}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.9\textwidth]{sd-approx-BRCA}}
\caption{Standard deviations derived from summary statistics of breast cancer based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set. Coloring shows the quality control applied in this paper.}
\label{fig:sd-approx-brca}
\end{figure}
